---
title: "F1 Visualizations"
description: |
  A few graphics for visualizing F1 data and a short script for harvesting current points data from the F1 website.
author:
  - name: Eduard Schlenz
    url: 
    affiliation: 
    affiliation_url:
date: "`r Sys.Date()`"
output: distill::distill_article
---
# Introduction
This analysis explores the performance of Formula 1 teams and drivers from 1994 to 2025, visualizing their accumulated wins, wins per season, and the winning teams in each race. We will also explore trends in team dominance and seasonal variations in wins across different periods.

```{r, message = FALSE}
# Required libraries for data manipulation and visualization
library(tidyverse)
library(lubridate)
library(jsonlite)
library(rvest)
library(dplyr)
library(ggplot2)
library(purrr)
library(httr)
library(xml2)
library(RColorBrewer)
library(knitr)
```

# Read Data
We begin by accessing the F1 race results using an API that provides historical results. The data includes season, race results, driver names, and constructor (team) information. We will collect this data to analyze the performance of teams and drivers.

```{r, message = FALSE}
# URL to fetch F1 race results from about 1994 to today
results_url <- "https://api.jolpi.ca/ergast/f1/results.json?limit=100&offset=0"

# Empty list to fill with content from ergast
ergast_content <- list()

# Loop to fetch data for all available seasons
for (offset in seq(10000, 25400, by = 100)) {
  url <- paste0(results_url, offset)
  
  response <- GET(url)

   # Check if the request was successful
  if (status_code(response) == 200) {
    # Parse the JSON content
    data <- fromJSON(content(response, "text"))
    
    # Append the data to the list
    ergast_content <- append(ergast_content, list(data))
    
  } else {
    # Handle error
    print(paste("Error: ", status_code(response)))
  }
    # 0.05 Second delay to stay within API limit
  Sys.sleep(0.05)
} 
# Takes about 1 minute to run
```

# Data Preparation: Race Results
In this section, we extract the race results, including the winning drivers and teams, and filter out incomplete data. We focus on the driver and constructor information to analyze team and driver performance.

```{r}

# Extract race results and drivers performance
races_list <- list()

# Extract the Drivers data frame from each element in all_drivers[[1]] to all_drivers[[9]]
for (i in 1:155) {
  races_list[[i]] <- ergast_content[[i]][["MRData"]][["RaceTable"]][["Races"]]
}
races_df <- bind_rows(races_list)

# Extract winning driver and constructor information
results_list <- races_df[["Results"]]

winnersd <- sapply(results_list, function(df) {
  winner_row <- df[df$position == 1, ]
  if (nrow(winner_row) > 0) {
    return(winner_row$Driver$driverId)
  } else {
    return("NA")
  }
})

winnerst <- sapply(results_list, function(df) {
  winner_row <- df[df$position == 1, ]
  if (nrow(winner_row) > 0) {
    return(winner_row$Constructor$constructorId)
  } else {
    return("NA")
  }
})

# Combine winner data into a data frame
winnersd <- data.frame(driver = winnersd)
winnerst <- data.frame(team = winnerst) 

driver_df <- bind_cols(winnersd,winnerst)

# Combine driver and team information with race data
results_df_raw <- bind_cols(driver_df, races_df)

# Filter out invalid data
results_df_raw <- results_df_raw %>%
  filter(!grepl("NA", driver))
```

# Data Preparation: Cleaned Data Frame
Now, we clean the data by selecting only the relevant columns (driver, season, team, and race name) for further analysis.

```{r}
# Selecting only the relevant columns for analysis
results_df <- results_df_raw %>%
  select(driver, season, team, raceName)
```

# Visualization1: Stair Step Graph of Wins Over Time
In this section, we visualize the cumulative wins for each team across all seasons using a stair-step graph. This graph highlights how the teams' performance has evolved over the years.

```{r}
# Accumulating team wins per season and visualizing using a stair-step graph
results_wins_acc <- results_df %>%
  group_by(season, team) %>%
  count(team)

# Adding cumulative wins for each team
results_wins_acc <- results_wins_acc %>%
  group_by(team) %>%
  mutate(cum_n = cumsum(n))

results_wins_acc$season <- as.numeric(results_wins_acc$season)

team_list <- c("ferrari", "mclaren", "mercedes", "red_bull", "williams", "renault", "benetton", "jordan", "brawn", "stewart", "alphatauri")

custom_team_colors <- c(
  "ferrari"   = "#DC0000",  
  "mclaren"   = "#FF8700",   
  "mercedes"  = "#00D2BE",   
  "red_bull"  = "#1E41FF", 
  "williams"  = "#005AFF",  
  "renault"   = "black",   
  "benetton"  = "#00B140",   
  "jordan"     = "#FFF500"  
)

results_wins_acc2 <- results_wins_acc %>%
  filter(team %in% team_list)

# Plotting the stair-step graph
h = ggplot(results_wins_acc2, aes(x = season, y = cum_n, group = team, color = team)) +
  geom_step(size = 1) +
  theme_bw() +
  theme(aspect.ratio = 4/3, axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  scale_x_continuous(breaks = seq(1990, max(results_wins_acc$season), by = 5),
                     labels = seq(1990, max(results_wins_acc$season), by = 5)) +
  scale_color_manual(values = custom_team_colors) +
  labs(x = "Season", y = "Accumulated Wins", color = "Teams")
h

```
# Visualization2: Wins per Season
Here, we show the number of wins per season for each team using a bar chart. This helps us to compare the performance of teams in different seasons.

```{r, layout="l-page-outset", fig.width=12, fig.height=7}
# Visualizing wins per season for each team
custom_team_colors <- c(
  "benetton"     = "darkgreen",
  "mclaren"      = "#FF8700",
  "williams"     = "purple",
  "ferrari"      = "#DC0000",
  "team_lotus"   = "gold",
  "ligier"       = "turquoise",
  "jordan"       = "#FFF500",
  "stewart"      = "grey",
  "renault"      = "black",
  "honda"        = "red",
  "bmw_sauber"   = "navy",
  "toro_rosso"   = "#1E41FF",
  "brawn"        = "limegreen",
  "red_bull"     = "#1E41FF",
  "lotus_f1"     = "gold",
  "mercedes"     = "#00D2BE",
  "alphatauri"   = "#1E41FF",
  "racing_point" = "pink",
  "alpine"       = "darkmagenta"
)
results_wins_acc3 <- results_wins_acc %>%
  filter(season >1986)

j = ggplot(results_wins_acc3, aes(x = as.factor(season), y = n, color = team, fill = team)) +
  geom_bar(stat = "identity", color = "white") +
  theme_bw() +
  theme(aspect.ratio = 4/3, 
        axis.text.x = element_text(angle = 90, size = 8, vjust = 0.35)) +
        scale_fill_manual(values = custom_team_colors) +
        labs(x = "Season", y = "Wins", fill = "Teams")
j
```

# Visualization3: Heatmap of Race Results
In this section, we visualize the winning teams for the top 25 races using a heatmap. The heatmap shows the distribution of winning teams across different seasons.

```{r, layout="l-page-outset", fig.width=12, fig.height=7}
# Preparing data for heatmap visualization
results_gp <- results_df %>%
  group_by(raceName) %>%
  summarise(Count = n())

# Transforming data for heatmap
results_wins_gp2 <- results_df %>%
  select(season, raceName, team) %>%
  spread(season, team)

# Joining with race count data
results_gp_joined <- results_wins_gp2 %>%
  left_join(results_gp, by = "raceName")

# Filtering top 25 races for the heatmap
results_gp_25 <- results_gp_joined %>%
  arrange(desc(Count)) %>%
  top_n(25, wt = Count) %>%
  slice(1:41)

# Pivoting data for heatmap
results_gp_25_long <- results_gp_25 %>%
  pivot_longer(cols = num_range("", 1994:2023),
               names_to = "season",
               values_to = "team")
# Top 10 teams for highlighting in the heatmap
team_counts <- results_gp_25_long %>%
  count(team, sort = TRUE)

top_10_teams <- team_counts %>%
  slice_max(n, n = 11)

# Adding category for top 10 teams
results_gp_25_long <- results_gp_25_long %>%
  mutate(team_category = ifelse(team %in% top_10_teams$team, as.character(team), "Others"))

custom_team_colors <- c(
  "ferrari"   = "#DC0000",   # Ferrari red
  "brawn"     = "limegreen",
  "mclaren"   = "#FF8700",   # Papaya orange (modern McLaren)
  "mercedes"  = "#00D2BE",   # Mercedes teal
  "red_bull"  = "#1E41FF",   # Red Bull blue
  "williams"  = "purple",   # Williams blue
  "renault"   = "black",   # Renault yellow
  "benetton"  = "darkgreen",   # Benetton green
  "jordan"    = "#FFF500",    # Honda red
  "lotus_f1"  = "gold",    # Honda red
  "Others"    = "lightgrey",
  "NA"        = "white"
)

# Heatmap plot
a = ggplot(results_gp_25_long, aes(x = season, y = raceName, fill = team_category)) +
  geom_tile(na.rm = TRUE, color = "black", position = "identity") +
  theme_classic() + 
  #scale_fill_brewer(palette = "Set3") +
  scale_fill_manual(values = custom_team_colors, na.value = "white") +
  theme(aspect.ratio = 4/3,
        axis.text.y = element_text(size = 8),
        axis.text.x = element_text(angle = 90, size = 8, vjust = 0.35),
        legend.position = "right",
        legend.box = "horizontal",
        legend.text = element_text(size = 8),
        legend.key.size = unit(1, "cm"),
        legend.key.width = unit(1, "cm"),
        axis.line = element_blank(),
        axis.ticks = element_blank()
            )  +
  labs(x = "Season", 
       y = "Race (Top 25)",
       fill = "Winning Team (Top 10)")
a
```

# Web Scraping: F1 Team Names and Points
In this section, we use rvest to scrape the names and points of F1 teams from the official website, then visualize the points of each team in a bar chart.

```{r}
# Read the F1 Teams page
f1_html <- read_html("https://www.formula1.com/en/teams.html")


# Scrape team names
teamnames <- f1_html %>%
  html_elements(".f1-heading") %>%
  html_text2()

indices <- seq(1, 50, by = 5)

teamnames <- teamnames[indices]

# Scrape points
points <- f1_html %>%
  html_elements(".f1-heading-wide") %>% 
  html_text2() %>%
  as.numeric()

indices <- seq(1, 20, by = 2)

points <- points[indices]

# Combine into dataframe
f1_df <- tibble(Team = teamnames, Points = points)

# Visualizing team points
u = ggplot(f1_df, aes(x = Points, y = reorder(Team, -Points))) +
  geom_bar(stat = "identity", fill = "steelblue", color = "black") +
  labs(x = "Points", y = "") +
  theme_classic() +
  theme(aspect.ratio = 4/3,
        )
u
```

# Conclusion
The visualizations provide insights into the performance of Formula 1 teams and drivers from the last 30 years. The stair-step graph and heatmap show the dominance of certain teams over time, while the bar chart helps highlight individual season performances. The scraped data reveals the current performance of teams based on their points in the latest season.

# Sources
F1 Race Results: API from https://api.jolpi.ca/ergast

F1 Team Information: https://www.formula1.com/en/teams.html


